BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-08-09 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
edits

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the generation of labels.
BarnebyLives rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

@@

Five records were appropriately flagged for issues with auto fill increment of the longitude value, and `r lat_flag` records were also auto-flagged for increases in latitude values. 
All flags were correct, and in several instances more errors were found in the rows following the flagged values.

@@

Thematically, BarnebyLives is set up to cover Western North America. 
However, the package supports the use of a 'domain' being drawn over any of the conterminous United States. 
Several of the attributes which it collects and displays on labels, relate to topics which more senior curators are interested in, i.e. the administrative information on Township Section and Range (or 'TRS'), but are considered less valuable in other geographic regions.  
Further several of the abiotic variables on which it acquires information on: slope, aspect, and geology, have long been considered prominent drivers of plant distributions in semi-arid and montane systems and is warranted on a label in these types of systems, whereas curators in other regions may find this information superfluous. 
Finally, it is plausible botanists in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the Western United States we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.

>>>>

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the generation of labels.
BarnebyLives rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the autofilling of coordinate and date columns. 
After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

@@

Five records were appropriately flagged for issues with autofills of the longitude value, and `r lat_flag` records were also flagged for autofilling increases in latitude values. 
All flags were correct, and in several instances more errors were found in the rows following the flagged values.

@@

Thematically, many of the variables which BarnebyLives collects information on are tailored to Western North America. 
However, the package supports the use of a 'domain' being drawn over any of the conterminous United States. 
Several of the attributes which it collects and displays on labels, relate to topics which more senior curators are interested in, i.e. the administrative information on Township Section and Range (or 'TRS'), but are considered less valuable in other geographic regions.  
Further several of the abiotic variables on which it acquires information on: slope, aspect, and geology, have long been considered prominent drivers of plant distributions in semi-arid and montane environments and are warranted on an herbarium label for specimens collected from these environments, whereas curators in other regions may find this information superfluous. 
Finally, it is plausible botanists in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the Western United States we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.
